### 2018-01-10 submission/manuscript.tex
rewrite background - add clearer picture of the confusion in current understanding

Targeting the identification of biomarkers within stool seems logical
since it offers an easy and cost-effective way to stratify risk and the
current gold standard for diagnosis, a colonoscopy, can be
time-consuming and is not risk free. Although stool represents an easy
and less invasive way to assess risk, it is not clear how reflective
this sample is to the community on the adenoma or carcinoma. Some
studies have tried to assess this in health and disease but are limited
by their sample size {[}12,13{]}. Sampling the tissue directly may
provide clearer answers but is not without limitations. Due to bowel
prep the communities left for sampling may not be reflective of the
resident microbiota, but rather a collection of what is able to keep
adhered to the mucosa. Additionally, these samples contain more host
DNA, potentially limiting the types of analysis that can be done. It is
well known that low biomass samples can be very difficult to work with
and results can end up being study dependent due to the randomness of
contamination {[}14{]}. Due to these many differences that could arise
between stool and tissue, one question our meta-analysis aims to answer
is whether there are consistent patterns that emerge across studies
regardless of whether they used stool or tissue samples.

This intense focus on identifying biomarkers has found strong
associations with resident mouth microbes as potential CRC-associated
microbes in the gastrointestinal (GI) tract {[}15--17{]}. The main
bacteria of interest, arising from this set of microbes, has been those
within the \emph{Fusobacterium} genus. Yet, the question remains as to
whether or not this is indeed the most important genera to be focusing
on, since many microbiota-based studies typically have identified a
collection of oral microbes rather than single species from a single
genera {[}16,17{]}. Based on this discrepancy, the second question we
can answer with this meta-analysis is if there is one dominant
CRC-associated genera that can be identified across studies.

The identification of microbial biomarkers has mostly dominated the
study of CRC and has had an unattended consequence of reducing the focus
on changes that occur within the underlying resident community. This has
been borne out by the majority of previous studies, within stool,
tissue, and the only meta-analysis investigating this area to date, that
focus predominately on biomarker identification. Yet, how these
CRC-associated microbes interact with their community may be an
important component to consider, and has not been investigated in great
detail by other meta-analyses. In response to this gap, our study aims
to answer whether there are consistent detectable community differences
as disease severity increases.

In comparison to the previous meta-analysis, this study significantly
increases the total stool samples investigated, examines differences
between stool and tissue microbiota in the context of CRC, and takes a
more community centric approach rather than a biomarker focused approach
to investigating commonalities across study for the microbiota and CRC.
Importantly, this community centric approach could provide valuable
insights into the importance of accounting for the community in CRC
disease not previously provided by earlier meta-analysis studies
{[}11{]}.

Using both feces (n = 1737) and tissue (492 samples from 350
individuals) totaling over 2229 total samples across 14 studies
{[}12,16--28{]} {[}Table 1 \& 2{]}, we expand both the breadth and scope
of the previous meta-analysis to investigate whether the bacterial
community is an important risk factor or if select members can
recapitulate the same information for both adenoma and carcinoma. To
accomplish this we first assessed whether bacterial diversity changes
throughout disease (control to adenoma to carcinoma) and if it results
in an increased relative risk (RR) for adenoma or carcinoma. We then
assessed what genera, if any, increase or decrease the RR of adenoma or
carcinoma. Next, using Random Forest models, we analyzed whether the
full community or only the top 5 increased and top 5 decreased RR genera
combined resulted in better model classification, based on the area
under the curve (AUC). Our results suggest that the community changes as
disease severity worsens and that this community is important for
disease classification. Finally, we also examined at what effect and
sample size the studies that were used were powered for and the sample
size needed to get to an 80\% power. Our power analysis suggests that
each individual study used was underpowered for detecting effect size
differences of 10\% or below between the case and control groups.

@@

\textbf{\emph{Obtaining Data Sets:}} The studies used for this
meta-analysis were identified through the review articles written by
Keku, \emph{et al.} and Vogtmann, \emph{et al.} {[}29,30{]} and
additional studies not mentioned in the reviews were obtained based on
the authors' knowledge of the literature. Studies that used tissue or
feces as their sample source for 454 or Illumina 16S rRNA gene
sequencing analysis were included and only data sets that had sequences
available for analysis were included. Some studies were excluded because
they did not have publicly available sequences or did not have metadata
in which the authors were able to share and were excluded. After these
filtering steps, the following studies remained: Ahn, \emph{et al.}
{[}26{]}, Baxter, \emph{et al.} {[}16{]}, Brim, \emph{et al.} {[}22{]},
Burns, \emph{et al.} {[}27{]}, Chen, \emph{et al.} {[}19{]}, Dejea,
\emph{et al.} {[}24{]}, Flemer, \emph{et al.} {[}12{]}, Geng, \emph{et
al.} {[}28{]}, Hale, \emph{et al.} {[}18{]}, Kostic, \emph{et al.}
{[}15{]}, Lu, \emph{et al.} {[}21{]}, Sanapareddy, \emph{et al.}
{[}25{]}, Wang, \emph{et al.} {[}20{]}, Weir, \emph{et al.} {[}23{]},
and Zeller, \emph{et al.} {[}17{]}. The Zackular {[}31{]} study was not
included because the 90 individuals analyzed within the study are
contained within the larger Baxter study {[}16{]}. After sequence
processing all the case samples for the Kostic study had 100 or less
sequences remaining and was not used which left a total of 14 studies
that analysis could be completed on.

@@

\textbf{\emph{Sequence Processing:}} For the majority of studies, raw
sequences were downloaded from the Sequence Read Archive (SRA)
(\url{ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/})
and metadata was obtained by searching the respective accession number
of the study at the following website:
\url{http://www.ncbi.nlm.nih.gov/Traces/study/}. Of the studies that did
not have sequences and metadata on the SRA, data was obtained from DBGap
(n = 1, {[}26{]}) and directly from the authors (n = 4,
{[}12,18,23,25{]}). Each study was processed using the mothur (v1.39.3)
software program {[}32{]} and quality filtering utilized the default
methods, used in mothur, for either 454 or Illumina based sequencing. If
it was not possible to use the defaults, the stated quality cut-offs,
from the study itself, were used instead. Chimeras were identified and
removed using VSEARCH {[}33{]} before \emph{de novo} OTU clustering at
97\% similarity was completed using the OptiClust algorithm {[}34{]}.

\textbf{\emph{Statistical Analysis:}} All statistical analysis after
sequence processing utilized the R (v3.4.3) software package {[}35{]}.
For the \(\alpha\)-diversity analysis, values were power transformed
using the rcompanion (v1.10.1) package {[}36{]} and then Z-score
normalized using the car (v2.1.5) package {[}37{]}. Testing for
\(\alpha\)-diversity differences utilized linear mixed-effect models
created using the lme4 (v1.1.14) package {[}38{]} to correct for study,
repeat sampling of individuals (tissue only), and 16S hypervariable
region used. Relative risk was analyzed using both the epiR (v0.9.87)
and metafor (v2.0.0) packages {[}39,40{]} by assessing how many
individuals with and without disease were above and below the overall
median value within each specific study. Relative risk significance
testing utilized the chi-squared test. \(\beta\)-diversity differences
utilized a Bray-Curtis distance matrix and PERMANOVA executed with the
vegan (v2.4.4) package {[}41{]}. Random Forest models were built using
both the caret (v6.0.77) and randomForest (v4.6.12) packages
{[}42,43{]}. Power analysis and estimations were made using the pwr
(v1.2.1) and statmod (v1.4.30) packages {[}44,45{]}. All figures were
created using both ggplot2 (v2.2.1) and gridExtra (v2.3) packages
{[}46,47{]}.

@@

\hypertarget{ref-flemer_tumour-associated_2017}{}
12. Flemer B, Lynch DB, Brown JMR, Jeffery IB, Ryan FJ, Claesson MJ, et
al. Tumour-associated and non-tumour-associated microbiota in colorectal
cancer. Gut. 2017;66:633--43.

\hypertarget{ref-Flynn_preprint_2017}{}
13. Flynn KJ, Ruffin MT, Turgeon DK, Schloss PD. Spatial variation of
the native colon microbiota in healthy adults. Cold Spring Harbor
Laboratory; 2017; Available from: \url{https://doi.org/10.1101/189886}

\hypertarget{ref-Salter_contamination_2014}{}
14. Salter SJ, Cox MJ, Turek EM, Calus ST, Cookson WO, Moffatt MF, et
al. Reagent and laboratory contamination can critically impact
sequence-based microbiome analyses. BMC Biology {[}Internet{]}. Springer
Nature; 2014;12. Available from:
\url{https://doi.org/10.1186/s12915-014-0087-z}

\hypertarget{ref-kostic_genomic_2012}{}
15. Kostic AD, Gevers D, Pedamallu CS, Michaud M, Duke F, Earl AM, et
al. Genomic analysis identifies association of Fusobacterium with
colorectal carcinoma. Genome Research. 2012;22:292--8.

\hypertarget{ref-baxter_microbiota-based_2016}{}
16. Baxter NT, Ruffin MT, Rogers MAM, Schloss PD. Microbiota-based model
improves the sensitivity of fecal immunochemical test for detecting
colonic lesions. Genome Medicine. 2016;8:37.

@@

\hypertarget{ref-hale_shifts_2017}{}
18. Hale VL, Chen J, Johnson S, Harrington SC, Yab TC, Smyrk TC, et al.
Shifts in the Fecal Microbiota Associated with Adenomatous Polyps.
Cancer Epidemiology, Biomarkers \& Prevention: A Publication of the
American Association for Cancer Research, Cosponsored by the American
Society of Preventive Oncology. 2017;26:85--94.

\hypertarget{ref-chen_human_2012}{}
19. Chen W, Liu F, Ling Z, Tong X, Xiang C. Human intestinal lumen and
mucosa-associated microbiota in patients with colorectal cancer. PloS
One. 2012;7:e39743.

\hypertarget{ref-wang_structural_2012}{}
20. Wang T, Cai G, Qiu Y, Fei N, Zhang M, Pang X, et al. Structural
segregation of gut microbiota between colorectal cancer patients and
healthy volunteers. The ISME journal. 2012;6:320--9.

\hypertarget{ref-lu_mucosal_2016}{}
21. Lu Y, Chen J, Zheng J, Hu G, Wang J, Huang C, et al. Mucosal
adherent bacterial dysbiosis in patients with colorectal adenomas.
Scientific Reports. 2016;6:26337.

\hypertarget{ref-brim_microbiome_2013}{}
22. Brim H, Yooseph S, Zoetendal EG, Lee E, Torralbo M, Laiyemo AO, et
al. Microbiome analysis of stool samples from African Americans with
colon polyps. PloS One. 2013;8:e81352.

\hypertarget{ref-weir_stool_2013}{}
23. Weir TL, Manter DK, Sheflin AM, Barnett BA, Heuberger AL, Ryan EP.
Stool microbiome and metabolome differences between colorectal cancer
patients and healthy adults. PloS One. 2013;8:e70803.

\hypertarget{ref-dejea_microbiota_2014}{}
24. Dejea CM, Wick EC, Hechenbleikner EM, White JR, Mark Welch JL,
Rossetti BJ, et al. Microbiota organization is a distinct feature of
proximal colorectal cancers. Proceedings of the National Academy of
Sciences of the United States of America. 2014;111:18321--6.

\hypertarget{ref-sanapareddy_increased_2012}{}
25. Sanapareddy N, Legge RM, Jovov B, McCoy A, Burcal L, Araujo-Perez F,
et al. Increased rectal microbial richness is associated with the
presence of colorectal adenomas in humans. The ISME journal.
2012;6:1858--68.

\hypertarget{ref-ahn_human_2013}{}
26. Ahn J, Sinha R, Pei Z, Dominianni C, Wu J, Shi J, et al. Human gut
microbiome and risk for colorectal cancer. Journal of the National
Cancer Institute. 2013;105:1907--11.

\hypertarget{ref-burns_virulence_2015}{}
27. Burns MB, Lynch J, Starr TK, Knights D, Blekhman R. Virulence genes
are a signature of the microbiome in the colorectal tumor
microenvironment. Genome Medicine. 2015;7:55.

\hypertarget{ref-geng_diversified_2013}{}
28. Geng J, Fan H, Tang X, Zhai H, Zhang Z. Diversified pattern of the
human colorectal cancer microbiome. Gut Pathogens. 2013;5:2.

\hypertarget{ref-keku_gastrointestinal_2015}{}
29. Keku TO, Dulal S, Deveaux A, Jovov B, Han X. The gastrointestinal
microbiota and colorectal cancer. American Journal of Physiology -
Gastrointestinal and Liver Physiology {[}Internet{]}. 2015 {[}cited 2017
Oct 30{]};308:G351--63. Available from:
\url{http://ajpgi.physiology.org/lookup/doi/10.1152/ajpgi.00360.2012}

\hypertarget{ref-vogtmann_epidemiologic_2016}{}
30. Vogtmann E, Goedert JJ. Epidemiologic studies of the human
microbiome and cancer. British Journal of Cancer {[}Internet{]}. 2016
{[}cited 2017 Oct 30{]};114:237--42. Available from:
\url{http://www.nature.com/doifinder/10.1038/bjc.2015.465}

\hypertarget{ref-zackular_human_2014}{}
31. Zackular JP, Rogers MAM, Ruffin MT, Schloss PD. The human gut
microbiome as a screening tool for colorectal cancer. Cancer Prevention
Research (Philadelphia, Pa.). 2014;7:1112--21.

\hypertarget{ref-schloss_introducing_2009}{}
32. Schloss PD, Westcott SL, Ryabin T, Hall JR, Hartmann M, Hollister
EB, et al. Introducing mothur: Open-Source, Platform-Independent,
Community-Supported Software for Describing and Comparing Microbial
Communities. Appl.Environ.Microbiol. {[}Internet{]}. 2009 {[}cited 12AD
Jan 1{]};75:7537--41. Available from:
\url{http://aem.asm.org/cgi/content/abstract/75/23/7537}

\hypertarget{ref-rognes_vsearch_2016}{}
33. Rognes T, Flouri T, Nichols B, Quince C, Mahé F. VSEARCH: A
versatile open source tool for metagenomics. PeerJ. 2016;4:e2584.

\hypertarget{ref-westcott_opticlust_2017}{}
34. Westcott SL, Schloss PD. OptiClust, an Improved Method for Assigning
Amplicon-Based Sequence Data to Operational Taxonomic Units. mSphere.
2017;2.

\hypertarget{ref-r_citation_2017}{}
35. R Core Team. R: A language and environment for statistical computing
{[}Internet{]}. Vienna, Austria: R Foundation for Statistical Computing;
2017. Available from: \url{https://www.R-project.org/}

\hypertarget{ref-rcompanion_citation_2017}{}
36. Mangiafico S. Rcompanion: Functions to support extension education
program evaluation {[}Internet{]}. 2017. Available from:
\url{https://CRAN.R-project.org/package=rcompanion}

\hypertarget{ref-car_citation_2011}{}
37. Fox J, Weisberg S. An R companion to applied regression
{[}Internet{]}. Second. Thousand Oaks CA: Sage; 2011. Available from:
\url{http://socserv.socsci.mcmaster.ca/jfox/Books/Companion}

\hypertarget{ref-lme4_citation_2015}{}
38. Bates D, Mächler M, Bolker B, Walker S. Fitting linear mixed-effects
models using lme4. Journal of Statistical Software. 2015;67:1--48.

\hypertarget{ref-epir_citation_2017}{}
39. Telmo Nunes MS with contributions from, Heuer C, Marshall J, Sanchez
J, Thornton R, Reiczigel J, et al. EpiR: Tools for the analysis of
epidemiological data {[}Internet{]}. 2017. Available from:
\url{https://CRAN.R-project.org/package=epiR}

\hypertarget{ref-metafor_citation_2010}{}
40. Viechtbauer W. Conducting meta-analyses in R with the metafor
package. Journal of Statistical Software {[}Internet{]}. 2010;36:1--48.
Available from: \url{http://www.jstatsoft.org/v36/i03/}

\hypertarget{ref-vegan_citation_2017}{}
41. Oksanen J, Blanchet FG, Friendly M, Kindt R, Legendre P, McGlinn D,
et al. Vegan: Community ecology package {[}Internet{]}. 2017. Available
from: \url{https://CRAN.R-project.org/package=vegan}

\hypertarget{ref-caret_citation_2017}{}
42. Jed Wing MKC from, Weston S, Williams A, Keefer C, Engelhardt A,
Cooper T, et al. Caret: Classification and regression training
{[}Internet{]}. 2017. Available from:
\url{https://CRAN.R-project.org/package=caret}

\hypertarget{ref-randomforest_citation_2002}{}
43. Liaw A, Wiener M. Classification and regression by randomForest. R
News {[}Internet{]}. 2002;2:18--22. Available from:
\url{http://CRAN.R-project.org/doc/Rnews/}

\hypertarget{ref-pwr_citation_2017}{}
44. Champely S. Pwr: Basic functions for power analysis {[}Internet{]}.
2017. Available from: \url{https://CRAN.R-project.org/package=pwr}

\hypertarget{ref-statmod_citation_2016}{}
45. Giner G, Smyth GK. Statmod: Probability calculations for the inverse
gaussian distribution. R Journal. 2016;8:339--51.

\hypertarget{ref-ggplot2_citation_2009}{}
46. Wickham H. Ggplot2: Elegant graphics for data analysis
{[}Internet{]}. Springer-Verlag New York; 2009. Available from:
\url{http://ggplot2.org}

>>>>

Although there has been an intense focus on microbiota based biomarker
discovery for CRC, the candidate genera seem to be numerous. Some
studies point towards mouth-associated genera such as
\emph{Fusobacterium}, \emph{Peptostreptococcus}, \emph{Parvimonas}, and
\emph{Porphyromonas} as key enriched genera {[}6,12--18{]}. Yet, even in
some of these studies mouth-associated genera are far from the only
microbes identified to be associated with CRC. These other genera
include, but are not necessarily limited to, \emph{Providencia},
\emph{Mogibacterium}, \emph{Enterococcus}, \emph{Escherichia/Shigella},
\emph{Klebsiella}, and \emph{Streptococcus} {[}14--16{]}. In fact,
\emph{Escherichia/Shigella} also has good \emph{in vivo} evidence that
it could be important in the pathogenesis of CRC {[}5,19{]}. Other
studies have also identified \emph{Akkermansia muciniphila} and
\emph{Bacteroides fragilis} as potential additional markers of CRC with
\emph{Bacteroides fragilis} having good mechanistic studies into how
this may occur {[}15,20,21{]}. Still, other studies have identified
genera like \emph{Roseburia} as increased in CRC while others have
reported it to be decreased or have found no difference in any genera at
all {[}15,18,22,23{]}.

Most of these studies have focused on carcinoma but the adenoma results
are not any clearer at identifying candidate genera. Some groups have
focused on more broad scale community metrics and have found that
metrics such as richness had decreased in adenoma versus controls while
others have identified \emph{Lactococcus}, \emph{Pseudomonas},
\emph{Acidovorax}, \emph{Cloacibacterium}, \emph{Helicobacter},
\emph{Lactobacillus}, \emph{Bilophila}, \emph{Desulfovibrio}, and
\emph{Mogibacterium} to be increased in adenoma {[}24--26{]}. A recent
meta-analysis confirmed the associations of mouth-associated genera and
\emph{Akkermansia muciniphila} in carcinoma {[}11{]}. However, it only
analyzed a total of 509 stool samples and with some recent studies, that
were not included, matching that sample size alone it is hard to know
how extrapolatable these findings truly are. This meta-analysis also
added more potential microbial associations to both carcinoma
(\emph{Pantoea agglomerans} \emph{Ruminococcus}, \emph{Lactobacillus})
and adenoma (\emph{Prevotella}, \emph{Methanosphaera},
\emph{Succinovibrio}, \emph{Haemophilus parainfluenzae},
\emph{Ruminococcus}, \emph{Lactobacillus}) that need to be investigated
further since a number of these genera have been found to be enriched in
controls and not lesion (adenoma or carcinoma) {[}13,16,17{]}.

Targeting the identification of biomarkers within stool seems logical
since it offers an easy and cost-effective way to stratify risk and the
current gold standard for diagnosis, a colonoscopy, can be
time-consuming and is not risk free. Although stool represents an easy
and less invasive way to assess risk, it is not clear how reflective
this sample is to the community on the adenoma or carcinoma. Some
studies have tried to assess this in health and disease but are limited
by their sample size {[}18,27{]}. Sampling the tissue directly may
provide clearer answers but is not without limitations. Due to bowel
prep the communities left for sampling may not be reflective of the
resident microbiota, but rather a collection of what is able to keep
adhered to the mucosa. Additionally, these samples contain more host
DNA, potentially limiting the types of analysis that can be done. It is
well known that low biomass samples can be very difficult to work with
and results can end up being study dependent due to the randomness of
contamination {[}28{]}. Due to these many differences that could arise
between stool and tissue, one question our meta-analysis aims to answer
is whether there are consistent patterns that emerge across studies
regardless of whether they used stool or tissue samples.

In comparison to the previous meta-analysis, this study significantly
increases the total stool samples investigated, re-examines important
genera across adenoma and carcinoma across study, and examines
differences and similarities between stool and tissue microbiota in the
context of CRC. Importantly, this analysis and approach could provide
valuable insights into the common genera that are both protective and
detrimental in CRC and whether broad bacterial community measurements
can account for these changes that were not previously provided by
earlier meta-analysis studies {[}11{]}.

Using both feces (n = 1737) and tissue (492 samples from 350
individuals) totaling over 2229 total samples across 14 studies
{[}12--18,20,22--26,29{]} {[}Table 1 \& 2{]}, we expand both the breadth
and scope of the previous meta-analysis to investigate whether the
bacterial community is an important risk factor or if select members can
recapitulate the same information for both adenoma and carcinoma. To
accomplish this we first assessed whether bacterial diversity changes
throughout disease (control to adenoma to carcinoma) and if it results
in an increased relative risk (RR) for adenoma or carcinoma. We then
assessed what genera, if any, increase or decrease the RR of adenoma or
carcinoma. Next, using Random Forest models, we analyzed whether the
full community or only the top 5 increased and top 5 decreased RR genera
combined resulted in better model classification, based on the area
under the curve (AUC). Our results suggest that the community changes as
disease severity worsens and that this community is important for
disease classification. Finally, we also examined at what effect and
sample size the studies that were used were powered for and the sample
size needed to get to an 80\% power. Our power analysis suggests that
each individual study used was underpowered for detecting effect size
differences of 10\% or below between the case and control groups.

@@

\textbf{\emph{Obtaining Data Sets:}} The studies used for this
meta-analysis were identified through the review articles written by
Keku, \emph{et al.} and Vogtmann, \emph{et al.} {[}30,31{]} and
additional studies not mentioned in the reviews were obtained based on
the authors' knowledge of the literature. Studies that used tissue or
feces as their sample source for 454 or Illumina 16S rRNA gene
sequencing analysis were included and only data sets that had sequences
available for analysis were included. Some studies were excluded because
they did not have publicly available sequences or did not have metadata
in which the authors were able to share and were excluded. After these
filtering steps, the following studies remained: Ahn, \emph{et al.}
{[}12{]}, Baxter, \emph{et al.} {[}13{]}, Brim, \emph{et al.} {[}29{]},
Burns, \emph{et al.} {[}16{]}, Chen, \emph{et al.} {[}14{]}, Dejea,
\emph{et al.} {[}23{]}, Flemer, \emph{et al.} {[}18{]}, Geng, \emph{et
al.} {[}22{]}, Hale, \emph{et al.} {[}26{]}, Kostic, \emph{et al.}
{[}32{]}, Lu, \emph{et al.} {[}25{]}, Sanapareddy, \emph{et al.}
{[}24{]}, Wang, \emph{et al.} {[}15{]}, Weir, \emph{et al.} {[}20{]},
and Zeller, \emph{et al.} {[}17{]}. The Zackular {[}33{]} study was not
included because the 90 individuals analyzed within the study are
contained within the larger Baxter study {[}13{]}. After sequence
processing all the case samples for the Kostic study had 100 or less
sequences remaining and was not used which left a total of 14 studies
that analysis could be completed on.

@@

\textbf{\emph{Sequence Processing:}} For the majority of studies, raw
sequences were downloaded from the Sequence Read Archive (SRA)
(\url{ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/})
and metadata was obtained by searching the respective accession number
of the study at the following website:
\url{http://www.ncbi.nlm.nih.gov/Traces/study/}. Of the studies that did
not have sequences and metadata on the SRA, data was obtained from DBGap
(n = 1, {[}12{]}) and directly from the authors (n = 4,
{[}18,20,24,26{]}). Each study was processed using the mothur (v1.39.3)
software program {[}34{]} and quality filtering utilized the default
methods, used in mothur, for either 454 or Illumina based sequencing. If
it was not possible to use the defaults, the stated quality cut-offs,
from the study itself, were used instead. Chimeras were identified and
removed using VSEARCH {[}35{]} before \emph{de novo} OTU clustering at
97\% similarity was completed using the OptiClust algorithm {[}36{]}.

\textbf{\emph{Statistical Analysis:}} All statistical analysis after
sequence processing utilized the R (v3.4.3) software package {[}37{]}.
For the \(\alpha\)-diversity analysis, values were power transformed
using the rcompanion (v1.10.1) package {[}38{]} and then Z-score
normalized using the car (v2.1.5) package {[}39{]}. Testing for
\(\alpha\)-diversity differences utilized linear mixed-effect models
created using the lme4 (v1.1.14) package {[}40{]} to correct for study,
repeat sampling of individuals (tissue only), and 16S hypervariable
region used. Relative risk was analyzed using both the epiR (v0.9.87)
and metafor (v2.0.0) packages {[}41,42{]} by assessing how many
individuals with and without disease were above and below the overall
median value within each specific study. Relative risk significance
testing utilized the chi-squared test. \(\beta\)-diversity differences
utilized a Bray-Curtis distance matrix and PERMANOVA executed with the
vegan (v2.4.4) package {[}43{]}. Random Forest models were built using
both the caret (v6.0.77) and randomForest (v4.6.12) packages
{[}44,45{]}. Power analysis and estimations were made using the pwr
(v1.2.1) and statmod (v1.4.30) packages {[}46,47{]}. All figures were
created using both ggplot2 (v2.2.1) and gridExtra (v2.3) packages
{[}48,49{]}.

@@

\hypertarget{ref-ahn_human_2013}{}
12. Ahn J, Sinha R, Pei Z, Dominianni C, Wu J, Shi J, et al. Human gut
microbiome and risk for colorectal cancer. Journal of the National
Cancer Institute. 2013;105:1907--11.

\hypertarget{ref-baxter_microbiota-based_2016}{}
13. Baxter NT, Ruffin MT, Rogers MAM, Schloss PD. Microbiota-based model
improves the sensitivity of fecal immunochemical test for detecting
colonic lesions. Genome Medicine. 2016;8:37.

\hypertarget{ref-chen_human_2012}{}
14. Chen W, Liu F, Ling Z, Tong X, Xiang C. Human intestinal lumen and
mucosa-associated microbiota in patients with colorectal cancer. PloS
One. 2012;7:e39743.

\hypertarget{ref-wang_structural_2012}{}
15. Wang T, Cai G, Qiu Y, Fei N, Zhang M, Pang X, et al. Structural
segregation of gut microbiota between colorectal cancer patients and
healthy volunteers. The ISME journal. 2012;6:320--9.

\hypertarget{ref-burns_virulence_2015}{}
16. Burns MB, Lynch J, Starr TK, Knights D, Blekhman R. Virulence genes
are a signature of the microbiome in the colorectal tumor
microenvironment. Genome Medicine. 2015;7:55.

@@

\hypertarget{ref-flemer_tumour-associated_2017}{}
18. Flemer B, Lynch DB, Brown JMR, Jeffery IB, Ryan FJ, Claesson MJ, et
al. Tumour-associated and non-tumour-associated microbiota in colorectal
cancer. Gut. 2017;66:633--43.

\hypertarget{ref-ecoli_Arthur_2014}{}
19. Arthur JC, Gharaibeh RZ, Mühlbauer M, Perez-Chanona E, Uronis JM,
McCafferty J, et al. Microbial genomic analysis reveals the essential
role of inflammation in bacteria-induced colorectal cancer. Nature
Communications {[}Internet{]}. Springer Nature; 2014;5:4724. Available
from: \url{https://doi.org/10.1038/ncomms5724}

\hypertarget{ref-weir_stool_2013}{}
20. Weir TL, Manter DK, Sheflin AM, Barnett BA, Heuberger AL, Ryan EP.
Stool microbiome and metabolome differences between colorectal cancer
patients and healthy adults. PloS One. 2013;8:e70803.

\hypertarget{ref-bfrag_Boleij_2014}{}
21. Boleij A, Hechenbleikner EM, Goodwin AC, Badani R, Stein EM, Lazarev
MG, et al. The bacteroides fragilis toxin gene is prevalent in the colon
mucosa of colorectal cancer patients. Clinical Infectious Diseases
{[}Internet{]}. Oxford University Press (OUP); 2014;60:208--15.
Available from: \url{https://doi.org/10.1093/cid/ciu787}

\hypertarget{ref-geng_diversified_2013}{}
22. Geng J, Fan H, Tang X, Zhai H, Zhang Z. Diversified pattern of the
human colorectal cancer microbiome. Gut Pathogens. 2013;5:2.

\hypertarget{ref-dejea_microbiota_2014}{}
23. Dejea CM, Wick EC, Hechenbleikner EM, White JR, Mark Welch JL,
Rossetti BJ, et al. Microbiota organization is a distinct feature of
proximal colorectal cancers. Proceedings of the National Academy of
Sciences of the United States of America. 2014;111:18321--6.

\hypertarget{ref-sanapareddy_increased_2012}{}
24. Sanapareddy N, Legge RM, Jovov B, McCoy A, Burcal L, Araujo-Perez F,
et al. Increased rectal microbial richness is associated with the
presence of colorectal adenomas in humans. The ISME journal.
2012;6:1858--68.

\hypertarget{ref-lu_mucosal_2016}{}
25. Lu Y, Chen J, Zheng J, Hu G, Wang J, Huang C, et al. Mucosal
adherent bacterial dysbiosis in patients with colorectal adenomas.
Scientific Reports. 2016;6:26337.

\hypertarget{ref-hale_shifts_2017}{}
26. Hale VL, Chen J, Johnson S, Harrington SC, Yab TC, Smyrk TC, et al.
Shifts in the Fecal Microbiota Associated with Adenomatous Polyps.
Cancer Epidemiology, Biomarkers \& Prevention: A Publication of the
American Association for Cancer Research, Cosponsored by the American
Society of Preventive Oncology. 2017;26:85--94.

\hypertarget{ref-Flynn_preprint_2017}{}
27. Flynn KJ, Ruffin MT, Turgeon DK, Schloss PD. Spatial variation of
the native colon microbiota in healthy adults. Cold Spring Harbor
Laboratory; 2017; Available from: \url{https://doi.org/10.1101/189886}

\hypertarget{ref-Salter_contamination_2014}{}
28. Salter SJ, Cox MJ, Turek EM, Calus ST, Cookson WO, Moffatt MF, et
al. Reagent and laboratory contamination can critically impact
sequence-based microbiome analyses. BMC Biology {[}Internet{]}. Springer
Nature; 2014;12. Available from:
\url{https://doi.org/10.1186/s12915-014-0087-z}

\hypertarget{ref-brim_microbiome_2013}{}
29. Brim H, Yooseph S, Zoetendal EG, Lee E, Torralbo M, Laiyemo AO, et
al. Microbiome analysis of stool samples from African Americans with
colon polyps. PloS One. 2013;8:e81352.

\hypertarget{ref-keku_gastrointestinal_2015}{}
30. Keku TO, Dulal S, Deveaux A, Jovov B, Han X. The gastrointestinal
microbiota and colorectal cancer. American Journal of Physiology -
Gastrointestinal and Liver Physiology {[}Internet{]}. 2015 {[}cited 2017
Oct 30{]};308:G351--63. Available from:
\url{http://ajpgi.physiology.org/lookup/doi/10.1152/ajpgi.00360.2012}

\hypertarget{ref-vogtmann_epidemiologic_2016}{}
31. Vogtmann E, Goedert JJ. Epidemiologic studies of the human
microbiome and cancer. British Journal of Cancer {[}Internet{]}. 2016
{[}cited 2017 Oct 30{]};114:237--42. Available from:
\url{http://www.nature.com/doifinder/10.1038/bjc.2015.465}

\hypertarget{ref-kostic_genomic_2012}{}
32. Kostic AD, Gevers D, Pedamallu CS, Michaud M, Duke F, Earl AM, et
al. Genomic analysis identifies association of Fusobacterium with
colorectal carcinoma. Genome Research. 2012;22:292--8.

\hypertarget{ref-zackular_human_2014}{}
33. Zackular JP, Rogers MAM, Ruffin MT, Schloss PD. The human gut
microbiome as a screening tool for colorectal cancer. Cancer Prevention
Research (Philadelphia, Pa.). 2014;7:1112--21.

\hypertarget{ref-schloss_introducing_2009}{}
34. Schloss PD, Westcott SL, Ryabin T, Hall JR, Hartmann M, Hollister
EB, et al. Introducing mothur: Open-Source, Platform-Independent,
Community-Supported Software for Describing and Comparing Microbial
Communities. Appl.Environ.Microbiol. {[}Internet{]}. 2009 {[}cited 12AD
Jan 1{]};75:7537--41. Available from:
\url{http://aem.asm.org/cgi/content/abstract/75/23/7537}

\hypertarget{ref-rognes_vsearch_2016}{}
35. Rognes T, Flouri T, Nichols B, Quince C, Mahé F. VSEARCH: A
versatile open source tool for metagenomics. PeerJ. 2016;4:e2584.

\hypertarget{ref-westcott_opticlust_2017}{}
36. Westcott SL, Schloss PD. OptiClust, an Improved Method for Assigning
Amplicon-Based Sequence Data to Operational Taxonomic Units. mSphere.
2017;2.

\hypertarget{ref-r_citation_2017}{}
37. R Core Team. R: A language and environment for statistical computing
{[}Internet{]}. Vienna, Austria: R Foundation for Statistical Computing;
2017. Available from: \url{https://www.R-project.org/}

\hypertarget{ref-rcompanion_citation_2017}{}
38. Mangiafico S. Rcompanion: Functions to support extension education
program evaluation {[}Internet{]}. 2017. Available from:
\url{https://CRAN.R-project.org/package=rcompanion}

\hypertarget{ref-car_citation_2011}{}
39. Fox J, Weisberg S. An R companion to applied regression
{[}Internet{]}. Second. Thousand Oaks CA: Sage; 2011. Available from:
\url{http://socserv.socsci.mcmaster.ca/jfox/Books/Companion}

\hypertarget{ref-lme4_citation_2015}{}
40. Bates D, Mächler M, Bolker B, Walker S. Fitting linear mixed-effects
models using lme4. Journal of Statistical Software. 2015;67:1--48.

\hypertarget{ref-epir_citation_2017}{}
41. Telmo Nunes MS with contributions from, Heuer C, Marshall J, Sanchez
J, Thornton R, Reiczigel J, et al. EpiR: Tools for the analysis of
epidemiological data {[}Internet{]}. 2017. Available from:
\url{https://CRAN.R-project.org/package=epiR}

\hypertarget{ref-metafor_citation_2010}{}
42. Viechtbauer W. Conducting meta-analyses in R with the metafor
package. Journal of Statistical Software {[}Internet{]}. 2010;36:1--48.
Available from: \url{http://www.jstatsoft.org/v36/i03/}

\hypertarget{ref-vegan_citation_2017}{}
43. Oksanen J, Blanchet FG, Friendly M, Kindt R, Legendre P, McGlinn D,
et al. Vegan: Community ecology package {[}Internet{]}. 2017. Available
from: \url{https://CRAN.R-project.org/package=vegan}

\hypertarget{ref-caret_citation_2017}{}
44. Jed Wing MKC from, Weston S, Williams A, Keefer C, Engelhardt A,
Cooper T, et al. Caret: Classification and regression training
{[}Internet{]}. 2017. Available from:
\url{https://CRAN.R-project.org/package=caret}
